Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-10-18 revision/r2/response_r2.tex

\externaldocument{forecastflows_r2}
\newcommand{\lref}[1]{}
% (\href{file:forecastflows_r1\#lintarget:SOMELABEL}{line \ref*{lin:SOMELABEL}})

% \DeclareEmphSequence{\itshape\color{darkgrey}}
\usepackage[most]{tcolorbox}
% \newtcolorbox{mybox}[1][]{%
%     blanker, 
%     left skip=0em,
%     left=1em,
%     borderline west={1pt}{2pt}{darkgrey},
%     breakable,
%     #1}
\newtcolorbox{mybox}[1][]{%
    blanker, 
    left skip=0em,
    left=0em,
    borderline west={0pt}{0pt}{white},
    breakable,
    coltext=darkgrey
    #1}

\begin{document}
\setlength{\parindent}{0cm}
\setlength{\parskip}{7pt}

Editor and reviewer comments are in \emph{gray italics} (we provide below the full context of the reviewers 1-2 comments below, and extract the parts of reviewer 3's comments that required changes to this manuscript), while our responses are in regular text. \\

@@

We thank the reviewer for their positive comments regarding our manuscript. 
We have worked to correct these inconsistencies.

{\bf Reviewer 2}

\begin{mybox}
\emph{All of my major comments were sufficiently addressed in this revision of
the manuscript. I appreciate that the supplement is now referenced in
the main document (I missed it on my last round of review). I have a few
editorial comments for the main text and a few minor comments about the
supplement, but I feel this manuscript is almost ready for publication.}
\end{mybox}

We are glad that the reviewer found the manuscript has improved, and thank them for all their feedback to improve the manuscript.

\begin{mybox}
\emph{\textbf{Manuscript Minor comments:}\\
Abstract: potential typo in second-to-last sentence. Should ``where
forecasting a natural output" be ``where forecasting is a natural output"?}
\end{mybox}
% vvdm18sept: is really a typo? I thought it was just a way to say it. Lizzie? 
% emw18Oct: It sounds weird and the reviewer does not like it so I changed it.
Corrected.

@@

\begin{mybox}
\emph{A couple typos in Box e.g., ``calibrating submodes separately allows to
avoid" missing "us"?}
\end{mybox}

% emwSep22: I changed this in the text and here; your previous version needs an 'it' (makes it possible) but I wanted to remove 'fact' so adjusted more. 
Changed to: ``Calibrating submodels separately makes it easier to avoid the likely reality that [...]''

\begin{mybox}
\emph{\textbf{Supplement Minor comments:}\\
Although the supplement works through an example far simpler than the
types of research questions discussed in the main text, it serves as a
reasonable (and accessible) example of how the workflow could be
implemented. There is a brief discussion towards the end of the
supplement about the ``scalability" of the proposed workflow, but it
leaves something to be desired. Have the authors applied this workflow
(or pieces of it) to examples that are more complex that could be cited,
or is that future work?}
\end{mybox}

We appreciate this point, however, we wanted our workflow to be approachable. We specifically wanted to leave room for readers to think through how they would advance upon it (to whatever degree of complexity they see as biological relevant), thus, we have left it as is.  
% emwSep22:  I actually hate the idea that a model needs to be complex, but it seems common among ecologists reviewing our workflow papers .... I might just say: We disagree that if the model is not sufficiently complex it leaves `something to be desired.' As we want this workflow to be approachable and leave room for readers to think through how they would advance upon it (to whatever degree of complexity they see as biological relevant), we have left it as is.

@@

\begin{mybox}
\emph{Stan model not included so could not be assessed (model1\_nc2.stan), I
recommend sharing for transparency.}
\end{mybox}

Apologies that we forgot to share the Stan code. We have now added the code in the workflow example.

\begin{mybox}
\emph{Figure 15, unclear what the posterior predictive distribution is of,
this detail could be added for transparency. Could also add many
posterior predictive datasets to the same plot and then overlay the
observed data to facilitate easier comparisons between y-tilde and y-obs.}
\end{mybox}

We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations, and show nine (9) draws.
% Hmm I don't understand?
% emwSep22:  I think this must be the figure on pg 15 -- we show just one draw of the PPC (red), we could overlap more (like 10-30) in a lighter shade. And put the observed data black ON TOP so easier to see. I think this would be good to do.... just one draw of a PPC is okay, but not greatI IMHO.

@@

{\bf Reviewer 3} % emwSep22:  I think you should extract only the parts that require a change to the current text. I will read the below once you have that done and am happy to read a separate letter to MA.

\begin{mybox}
\emph{First, I would like to congratulate the authors for a very clear and well written manuscript. The main thesis of the authors is to outline modelling workflows and illustrate in broad brush strikes how to bridge the current gap in ecological forecasting with simulations of synthetic data among the several steps of model building (the pre-data/post-model third corner in their Fig. 1) to improve confidence in modelling, inference and prediction. This is an important contribution, one that echoes many early calls (e.g. Thomas 1997): our community in ecology, despite its beefing up with respect to quantitative skills, remains fragmented and disjoint. I thus wholeheartedly endorse the main thesis of the authors and agree with it about research. There remains one concern that I would like to discuss with the authors, and possibly hear back from them although it is somewhat minor with respect to research but more problematic with respect to policy. It should not deflect from publication of the article in Philosophical Transactions of the Royal Society A.}
\end{mybox}

We are glad that the reviewer found our manuscript well-written, and that he endorses our main messages. We are particularly grateful for his detailed comments, with much broader perspectives than just our manuscript. We believe that they deserve longer discussions (and could constitute in themselves another paper!), so we will reach out directly to M. Authier to share our detailed perspective on this. Below we address the comments that required changes to the current text.

@@

Thank you for this suggestion. We now use a darker gray to highlight this corner.

\begin{mybox}
\emph{Moving to the upper right corner (post-model/pre-data), the authors could mention
Simulation-Based Calibration (Modrák et al. 2025) to point readers to this useful tool for code
checking, which is also central to workflows. I suspect this omission might be wilful not to
overload readers. It might still be useful to point SBC on ecologists’ map as new models keep
being proposed but are insufficiently put to the test beyond one case study in many instances
(see also Nichols et al. 2019).}  
\end{mybox}

We now mention simulation-based calibration (\href{file:forecastflows_r2\#lintarget:SBC}{line \ref*{lin:SBC}}).

\iffalse
\begin{mybox}
\emph{With respect to terminology, the authors are mentioning “research spheres” in the abstract,
which is a catchy term, and one I agree may be more appropriate than that of a research
programme for what the authors are discussing. What is a research sphere is not articulated
further by the authors and there must be a “policy sphere” at which the paper is alluding to. In
fact, where I find the authors’ thesis is less convincing is with respect to policy outreach. In
policy, vague terminology such a “research sphere” is useful to get discussions started as
loosely defined terms facilitate interactions. This can also lead to talking past each other but,
usually, terminology gets to be clarified at some point when misunderstandings are too
numerous to be ignored (and one get a common understanding document out of a round of
meetings). My point is that the expectations of the “policy sphere” are usually not completely
aligned with the “research sphere”. And this realization bears on workflows.}  
\end{mybox}
\fi
...
% do we have t ochange something here?
% emw18Oct: I don't think so, I commented it out.

@@

% \begin{mybox}
% \emph{I would love to actually read the authors’ thoughts here and ideally be proven wrong that an
% improved workflow would be enough for enabling policy decisions whose beneficial impacts
% would be commensurate with the current magnitude of biodiversity loss7. In particular, I
% would be interested in a better articulation of how they see global studies feed-backing on
% policy which is usually more local, national or regional.}  
% \end{mybox}

% ...

\begin{mybox}
\emph{My remaining points are more concise.
- Terminology. If the authors are also aiming at the policy sphere, the word bias (e.g. lines
13, 134) should be wielded with extreme caution as it is easily weaponized. In policy
circles, the term bias is, in my experience, the kiss of death as it is often interpreted as
lopsided evidence. The authors are discussing preferential sampling with respect to space,
time or taxonomy and its consequences in modelling results. Preferential sampling is more
descriptive than bias, which also has a prescriptive meaning, esp. in policy. Also, a well-
designed survey at some spatial scale may become an instance of preferential geographic
sampling when collated with other surveys and data. This sort of data collation is often the
case with global trend analyses that the authors seem to have in mind.}  
\end{mybox}

This is a good point, we have clarified our use of this word in the two areas we used it without additional context ( \llabel{bias1}, \llabel{bias1}). % emw18Oct: fix these refs to main text!
% Hmm... well... I prefer bias. Preferential sampling sounds to me like it's done on purpose

@@

\begin{mybox}
\emph{Also, taxonomic biases are bound to happen in global analyses. How do we redress those? That is, what can
we say about data-poor species? How a richly-parametrized workflow calibrated on data
collected on a few very well-studied species is going to help the majority of data-poor
species? Taxonomic gaps have been identified without workflows and how to address
those gaps point to monitoring and the need of appropriate governance to ensure long-
term monitoring and inclusion of data-poor species. Research alone may not provide
enough incentive or impetus for inclusion of data-poor species, especially if what is
needed for policy about them is descriptive and routine science.}  
\end{mybox}

The workflow may allow to identify a missing phylogenetic structure that could be used to improve inference for rare species (with information from close taxa). Moreover, the workflow can also help identify the limits of our model---e.g. through data simulation, we could see that the model does not do a good job at predicting data-poor species. This could thus help to identify priorities for data collection.

\begin{mybox}
\emph{- ML enthusiasm (lines 298-303): the point is granted for prediction but again evidencing a
trend for policy is often required for further action. I fail to see how ML, if focused on
prediction, is going to help. A trend is a useful fiction that is easily understood in policy
fora where interpretability still has a huge premium. And a trend is very much
interpretable. I think the authors may be going overboard in their claims if, again, their
proposal is to have policy-relevance and not only research-relevance (a worthy goal of its
own).}  
\end{mybox}

% emw18Oct: friendly reminder to avoid 'sorry' in professional letters IMHO
Apologies this was not clear. Our point here was not to share any enthusiasm for machine learning, but rather to highlight the need for an even more careful and coherent workflow to build interpretable models (i.e. avoiding  ``complexity trap'' and build models with ``clear mechanistic drivers''). We tried to make this point more explicit  (\href{file:forecastflows_r2\#lintarget:clearerML}{line \ref*{lin:clearerML}}).

@@

\begin{mybox}
\emph{- typo on line 138: “These feedback loops are applied currently in only in Bayesian
frameworks (to our knowledge)...”}
\end{mybox}

Thank you for catching this! We fixed this typo.

\begin{mybox}
\emph{Last, I would like to re-iterate that I view this manuscript as a very useful contribution for the
research spheres in ecology. It can be published as is for this audience. I do not dispute any of
the claims made with respect to research, namely that disciplining model building to the
rationalizing force of workflows is a good thing. I dispute what I think are shortcuts and could
be perceived as a naïve outlook on the policy process. I do not assume that the authors
subscribe to the linear model of science but a too quick reading of their opinion piece may
suggest otherwise, especially the introduction when it refers to policy. For example, if lines 32-
41 are an accurate description of the research sphere, the policy sphere is then very rational in
keeping an instrumental outlook to trend detection and favouring simple methods until the
research sphere gets its workflows fully operational. The path from an improved workflow in
research to better policy decisions remains to be charted: scientific workflows are likely a
necessary step, yet one that may not be sufficient.}
\end{mybox}

We would like to thank the reviewer once again for all these very interesting avenues for discussion---and look forward to discussions with him directly. % beyond the community of researchers (to whom this paper was primarily addressed).

### 2025-10-18 revision/r2/forecastflows_r2.tex
Edits to letter and a couple more to main text

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about biodiversity loss and climate change have led to greater demands for ecological models. Datasets for developing these models have increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models. First, the research community is split between two spheres that prevent uniting trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, often including complex, mechanistic sub-models indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions. To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting a natural output. We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.

@@

Meeting these \llabel{global}global policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on fewer data. 
Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic biases. Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. These approaches have failed to yield agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.

@@

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see Supplementary Figure 1) or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.

>>>>

\begin{abstract}
% Concerns about increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Datasets relevant for developing these models and forecasts have also increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare most current models for several reasons. First, the research community is split between two disconnected paradigms: either using data to fit simple, trend-like models with one or few parameters, or developing forecasting models that include complex, mechanistic, and/or black box submodels only indirectly informed by data. Second, in both cases, models tend to be developed not only in isolation of one another, but also without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions, throughout the entire modeling process. We propose a unified, principled workflow for end-to-end empirical model development that bridges the gap between process-based and statistical approaches, integrates sound statistical and scientific practices, and especially relies on data simulation to inform decisions at multiple steps in the process. We argue that this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could transform for ecological modeling.
Concerns about biodiversity loss and climate change have led to greater demands for ecological models. Datasets for developing these models have increased in size and complexity, including in their geographical, temporal and phylogenetic dimensions. New research often suggests that models accounting for these complexities can yield more accurate trends and predictions. 
We argue, however, that the usual workflows for model fitting in ecology make it difficult to evaluate and compare current models. First, the research community is split between two spheres that prevent uniting trend estimation and forecasting. One research sphere focuses on using data to fit simple, trend-like models with few parameters, which is separated from research developing forecasting models, often including complex, mechanistic sub-models indirectly informed by data. Second, in both cases, models tend to be developed without a coherent framework for linking scientific questions and understanding to statistical and modeling decisions. To address these challenges we propose a workflow that integrates statistical and scientific practices through clear steps, many of which rely on data simulation to inform decisions in the process, and where forecasting is a natural output. We show how this approach, coupled with a shift toward universal training, more open model sharing, and alignment on common datasets, could harmonize currently divided efforts at trend estimation and forecasting to better inform sustainable policies.

@@

Meeting these \llabel{global}global policy needs has led often to two separate paths: one focused on estimating trends from new global datasets, and another focused on forecasting from generally distinct datasets or mechanistic models based on fewer data. 
Newly available large-scale, long-term datasets have provided our first `global' estimates of biodiversity trends \citep[e.g.][]{loh2005living,Dornelas2018}, but these data---gathered opportunistically from multiple sources---are unbalanced and suffer from large geographic, temporal and taxonomic sampling biases.\llabel{bias1} Models to date have failed to fully address these challenges and, perhaps because of these limitations, are rarely if ever used for forecasting.
Instead, forecasting---under different plausible scenarios---has generally relied on entirely different datasets combined with either correlative or process-based models \citep{IPBES2019}, with process-based models often promoted as the most realistic approach \citep{Urban2016, Pilowsky2022} because they focus on mechanistic representations of ecosystem functioning. These approaches have failed to yield agreement on current species trends, leading to ongoing debates about the magnitude and even direction \citep{Dornelas2014, Leung2020, Buschke2021, Johnson2024}, and producing forecasts that diverge due to high model uncertainty at the ecological level \citep{Cheaib2012, Thuiller2019}.

@@

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see Supplementary Figure 1) or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased (for example, more intensive sampling in Europe) \llabel{bias2}, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.
